Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2022-02-24 paper/paper.tex
Thursday work

\begin{figure}
\includegraphics[width=\linewidth]{./images/expressions_fig.png}
\caption{Simulated retinal and tectal signalling molecule expression. Four
receptor types are expressed, and four corresponding ligand types. Receptors
of type $i$ are activated by ligands of type $i$.
%
\textbf{A} Retinal receptor expression, $r_i$. Four receptors are expressed and
displayed here in two dual-colour maps. $r_0$ and $r_1$ are red and blue;
$r_2$ and $r_3$ are cyan and magenta. $r_0$ increases as an exponential
function of $-x$; $r_1$ increases with an exponential function of $-y$. $r_2$
and $r_3$ have the opposite sense. Retinal receptor expression is modelled
with exponential functions because there is ample evidence to support this
form. $N$: nasal, $T$: temporal, $D$: dorsal, $V$: ventral.
%
\textbf{B} Retinal ligand expression, $l_i$, is expressed in complimentary
gradients \citep{hornberger_modulation_1999}.
%
\textbf{C} Tectal ligand expression $L_i$. Tectal ligand expression is shown
in an exponential form,
although there is less evidence to determine whether tectal expression is
exponential, linear or logarithmic.
$L$: lateral, $M$: medial, $C$: caudal, $R$: rostral.
}
\label{f:ex}
\end{figure}

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_G.json configs/simpler/e_wt_fig2.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_G_wt_fig2.png}
\caption{The pure gradient-following model. \textbf{A} The origin position
on the tectum gives axons a colour according to a dual-colour map. Axons in
panels B-H that are red are those which originated in the most naso-ventral
corner of the square simulated retina. \textbf{B}-\textbf{D} Axon centroid
positions on the simulated tectum (dashed line) at time 0, 40 and 300
units. Lines show adjacency relationships between the axons' origin positions
on the retina. \textbf{E} Layout of axon centroids on the tectum observed in
experimental results. \textbf{F} The sum-of-squared difference error,
$\epsilon$, and the number-of-crossings error metric, $\eta$ of the tectal axon
arrangement plotted with respect to simulation time. \textbf{G} The
arrangement of axon branches at t=300. \textbf{H} Selected axon locations at
t=300, with path histories in grey.  \emph{Abbreviations: D dorsal, V ventral, T
temporal, N nasal, C caudal, R rostral, M medial, L lateral.}}
\label{f:G}
\end{figure}

@@

% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_wt_fig3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_G_wt_fig3.png}
\caption{By tuning the tectal ligand expression (replacing Eq.\,\ref{e:tecligexp} with
 Eq.\,\ref{e:tecligexp2}), the pure gradient-following model can achieve a
 successful layout in which the axons are arranged evenly across the full
 tectum. At the centre of the tectum, where neither retinal receptor
 expression nor tectal ligand expression is high, the ordering is weaker than
 towards the edge. \textbf{A} Axon branches \textbf{B} axon
 centroids \textbf{C} Five selected axons.}
\label{f:G0}
\end{figure}

@@

% The chemo manipulations figure is made up from 4 figures generated with
% graph_layout::g
%
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot90_fig4row1.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecrot180_fig4row2.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecswap_fig4row3.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_retablat_fig4row4.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_tecablat_fig4row5.json
% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_mismatch_fig4row6.json
%
% (see scripts/fig_G_surgical.sh)
\begin{figure}
\includegraphics[width=0.95\linewidth]{./images/fig_chemo_manipulations.png}
\caption{Surgical manipulations in the chemoaffinity-only model. \textbf{A} 90 degree
rotation. \textbf{B} 180 degree rotation. \textbf{C} Graft swap. \textbf{D}
Retinal ablation \textbf{E} Tectal ablation \textbf{F} mismatch ablation}
\label{f:Gsurg}
\end{figure}

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_GC_best_1.json configs/simpler/e_wt_figcomp1.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GC_best_1_wt_figcomp1.png}
\caption{$\mathbf{GC}$ model}
\label{f:GC}
\end{figure}

@@

Because the competition effect, $\mathbf{C}_b$ (Eq.\,\ref{e:C}), requires a
global supervisor to provide information to each branch about the location of
all other branches, we looked for a receptor-ligand interaction that could
signal between axon branches and enable them to adjust their movements away
from one another.

% First, look at I, from Simpson and Goodhill.
First, we took inspiration from Simpson and Goodhill's model for the
interaction between EphA receptors, which is defined (with receptor type $i$
standing for EphA) as:

@@

In order to replace the distance-based competition, $\mathbf{C}_b$, we first
suggest that each receptor ligand pair involved in the retinotectal system
could mediate a repulsive interaction. DO WE NEED TO? Taken alone, the EphA
interaction between one branch, $b$, and another, $k$, will cause a repulsion
if $k$ originated more \textbf{nasally?} than $b$. What about a branch
originating on the temporal side of $b$? This could be handled by the
interaction between $k$ and $b$ and may not require an additional interaction.

Next, we investigated the idea that the same receptor-ligand
interaction that leads to gradient-based movement of RGC axons on the tectum
could also act as a signal between the axons. This is plausible, because
retinal ganglion cells are known to express ephrin receptor ligands as well as
Eph receptors \citep{ephrins_on_RGCs}. We defined $N=4$ ligands, $l_i$,
expressed by the RGCs in the model according to
%
\begin{equation} \label{e:retlig}
l_i(x,y) = 1.05 + 0.26 \exp(2.3 u).
\end{equation}
%
We used exactly the same form as the receptor expression given by
Eq.\,\ref{e:retrcpt}, but with opposite gradient direction so here $u$ is
substituted by $x$ for $l_0$, $y$ for $l_1$, $(1-x)$ for $l_2$ and by $(1-y)$
for $l_3$ (see Fig.\,\ref{f:ex}C).
%
We combined chemotaxis with a multiplicative (or mass-action?) interaction,
$\mathbf{J}_b$ (scaled by a parameter $m_{\!_J}$) to define the movement
vector, $\mathbf{M}_b$, for the $\mathbf{GJ}$ model as:
%
\begin{equation} \label{e:mv3}
 \mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
 %%+  m_{\!_I} \mathbf{I}_b +  m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
%
$\mathbf{J}_b$ is defined as
%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{J}_b = \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,Q_{\!_J}(d_{kb})
\end{equation}
%
where $Q_{\!_J}(d_{kb})$ is the signalling strength between two growth cones of
receptor-ligand interaction radius $r_{\!_J}$ a distance $d_{kb}$ from one another,
given by
%
\begin{equation}
Q_{\!_J}(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r_{\!_J} \\
     0 & d_{kb} > 2r_{\!_J}
     \end{cases}
\end{equation}
%
The sign of $Q_{\!_J}$ (which is dependent on the values of $F_i$) determines
whether the interaction, $\mathbf{J}_b$, is repulsive or attractive. $l_{i,k}$
is the expression of ligand $i$ on branch $k$.

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_GI_best_1.json configs/simpler/e_wt_figcomp2.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GI_best_1_wt_figcomp2_exit_true.png}
\caption{$\mathbf{GI}$ model}
\label{f:GI}
\end{figure}

% ./scripts/fig_GI_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GI_surgical.png}
\caption{$\mathbf{GI}$ model, surgical manipulations}
\label{f:GIsurg}
\end{figure}

@@

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_GJ_best_1.json configs/simpler/e_wt_figcomp3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GJ_best_1_wt_figcomp3.png}
\caption{$\mathbf{GJ}$ model}
\label{f:GJ}
\end{figure}

>>>>

\begin{figure}
\includegraphics[width=\linewidth]{./images/expressions_fig.png}
\caption{Simulated retinal and tectal signalling molecule
expression. \textbf{A} A schematic of retino-tectal mapping.
%
\textbf{B} Retinal receptor expression, $r_i$. Four receptors are expressed and
displayed here in two dual-colour maps. $r_0$ and $r_1$ are red and blue;
$r_2$ and $r_3$ are cyan and magenta. $r_0$ increases as an exponential
function of $-x$; $r_1$ increases with an exponential function of $-y$. $r_2$
and $r_3$ have the opposite sense. Retinal receptor expression is modelled
with exponential functions because there is ample evidence to support this
form. $N$: nasal, $T$: temporal, $D$: dorsal, $V$: ventral.
%
\textbf{C} Retinal ligand expression, $l_i$, is expressed in complimentary
gradients \citep{hornberger_modulation_1999}.
%
\textbf{D} Tectal ligand expression $L_i$. Tectal ligand expression is shown
in an exponential form,
although there is less evidence to determine whether tectal expression is
exponential, linear or logarithmic.
$L$: lateral, $M$: medial, $C$: caudal, $R$: rostral.
}
\label{f:ex}
\end{figure}

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_G.json configs/simpler/e_wt_fig2.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_G_wt_fig2_exit_true.png}
\caption{The pure gradient-following model. \textbf{A} The origin position
on the tectum gives axons a colour according to a dual-colour map. Axons in
panels B-H that are red are those which originated in the most naso-ventral
corner of the square simulated retina. \textbf{B}-\textbf{D} Axon centroid
positions on the simulated tectum (dashed line) at time 0, 40 and 300
units. Lines show adjacency relationships between the axons' origin positions
on the retina. \textbf{E} Layout of axon centroids on the tectum observed in
experimental results. \textbf{F} The sum-of-squared difference error,
$\epsilon$, and the number-of-crossings error metric, $\eta$ of the tectal axon
arrangement plotted with respect to simulation time. \textbf{G} The
arrangement of axon branches at t=300. \textbf{H} Selected axon locations at
t=300, with path histories in grey.  \emph{Abbreviations: D dorsal, V ventral, T
temporal, N nasal, C caudal, R rostral, M medial, L lateral.}}
\label{f:G}
\end{figure}

@@

% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_wt_fig3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_G_wt_fig3_exit_true.png}
\caption{By tuning the tectal ligand expression (replacing Eq.\,\ref{e:tecligexp} with
 Eq.\,\ref{e:tecligexp2}), the pure gradient-following model can achieve a
 successful layout in which the axons are arranged evenly across the full
 tectum. At the centre of the tectum, where neither retinal receptor
 expression nor tectal ligand expression is high, the ordering is weaker than
 towards the edge. \textbf{A} Axon branches \textbf{B} axon
 centroids \textbf{C} Five selected axons.}
\label{f:G0}
\end{figure}

@@

% The chemo manipulations figure is made up from 4 figures generated with
% graph_layout::g
% (see scripts/fig_G_surgical.sh)
\begin{figure}
\includegraphics[width=0.95\linewidth]{./images/fig_chemo_manipulations.png}
\caption{Surgical manipulations in the chemoaffinity-only model. \textbf{A} 90 degree
rotation. \textbf{B} 180 degree rotation. \textbf{C} Graft swap. \textbf{D}
Retinal ablation \textbf{E} Tectal ablation \textbf{F} mismatch ablation}
\label{f:Gsurg}
\end{figure}

@@

@@

Because the competition effect, $\mathbf{C}_b$ (Eq.\,\ref{e:C}), requires a
global supervisor to provide information to each branch about the location of
all other branches (But \textbf{fixme}, arguably, so does the GJ model now!),
we looked for a receptor-ligand interaction that could signal between axon
branches and enable them to adjust their movements away from one another.

% First, look at I, from Simpson and Goodhill.
First, we took inspiration from Simpson and Goodhill's model for the
interaction between EphA receptors, which is defined (with receptor $r_i$
representing EphA) as:

@@

In order to replace the distance-based competition, $\mathbf{C}_b$, we first
point out that many of the ligands, $l_i$, expressed on the tectum are also
expressed by retinal ganglion cells and so where there is signalling between
tectal and RG cells, there is also the possibility for inter-RGC
signalling. Furthermore, if a receptor-ligand interaction on the tectum leads
to growth cones moving away from a region of high ligand expression, then the
equivalent axon-axon interaction would also be expected to be repulsive.

Taken alone, the EphA forward interaction between one branch, $b$, and
another, $k$, will cause a repulsion if $k$ originated more nasally than
$b$. A branch originating on the \emph{temporal} side of $b$ will cause no
repulsion of $b$. However, although $k$ does not induce repulsion in $b$, $b$
should induce a repulsion in $k$ and thus, it could be argued that only a
single ephrin-A ligand gradient should be required in the retina to ensure
that repulsive signals are transmitted between axons whose cell bodies
originte far apart in the retina. However, there is an asymmetry in the
interaction. Consider a cluster of growth cones, $k$, originating from the
nasal retina near to a single cone, $b$, from the temporal retina and assume
that $r_b/r_k>1.1$. The single cone is affected by the sum of 3 interactions
and has a movement of 3 units added to its movement vector (in the temporal
direction). Now reverse the originating locations so that the cluster, $k$,
originates temporally and the individual cone, $b$, comes from the nasal
retina. Now $r_k/r_b>1.1$ and each member of the cluster experiences an
addition of just one unit to its movement vector and the cones move apart at a
speed that is only $1/3$ of that for the original origins.

To avoid this asymmetry, we allow two receptor-ligand pairs, which are
expressed in opposing directions, to contribute to axon-axon interactions,
ensuring that interactions between single growth cones and clusters are not
dependent on their origin locations. Fig.\,\ref{f:expressions}C shows ligand
expressions representing an ephrin-A like expression ($l_0$) and an opposing
expression ($l_2$). It remains the case (with a suitable choice of threshold,
$s$) that branches originating from the same location on the temporal-nasal
axis will not experience competitive interactions.

Note that it would be possible to postulate two additional retinal ligand
expression fields (for $l_1$ and $l_3$), but we found this not to be
necessary; a single opposing pair of fields ensures that axons that originated
close together (and branches of the same axon) do not repel one another,
whereas axons that are close on the temporal-nasal axis, but far apart on the
dorso-ventral axis are generally driven far apart by chemotaxis, so whether
they compete or not when close together does not affect the behaviour of the
system.

% Now describe J model, and show the result.

% ./build/sim/agent/agent1 configs/simpler/m_ee_GJ_best_1.json configs/simpler/e_wt_figcomp3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GJ_best_1_wt_figcomp3_exit_true_steps_1500.png}
\caption{$\mathbf{GJ}$ model}
\label{f:GJ}
\end{figure}

%% TO ADD: Figure for the Reber/Brown manipulation on the GJ model.

\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GJ_best_1_reber.png}
\caption{$\mathbf{GJ}$ model, Reber et al.~manipulation. Look at THAT!}
\label{f:GJreber}
\end{figure}

%% HEREish
Next, we investigated the idea that the same receptor-ligand
interaction that leads to gradient-based movement of RGC axons on the tectum
could also act as a signal between the axons. This is plausible, because
retinal ganglion cells are known to express ephrin receptor ligands as well as
Eph receptors \citep{ephrins_on_RGCs}. We defined $N=4$ ligands, $l_i$,
expressed by the RGCs in the model according to
%
\begin{equation} \label{e:retlig}
l_i(x,y) = 1.05 + 0.26 \exp(2.3 u).
\end{equation}
%
We used exactly the same form as the receptor expression given by
Eq.\,\ref{e:retrcpt}, but with opposite gradient direction so here $u$ is
substituted by $x$ for $l_0$, $y$ for $l_1$, $(1-x)$ for $l_2$ and by $(1-y)$
for $l_3$ (see Fig.\,\ref{f:ex}C).
%
We combined chemotaxis with a multiplicative (or mass-action?) interaction,
$\mathbf{J}_b$ (scaled by a parameter $m_{\!_J}$) to define the movement
vector, $\mathbf{M}_b$, for the $\mathbf{GJ}$ model as:
%
\begin{equation} \label{e:mv3}
 \mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
 %%+  m_{\!_I} \mathbf{I}_b +  m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
%
$\mathbf{J}_b$ is defined as
%
\begin{equation}
% No \frac{1}{|B_b|} in this one. Really?
\mathbf{J}_b = \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,Q_{\!_J}(d_{kb})
\end{equation}
%
where $Q_{\!_J}(d_{kb})$ is the signalling strength between two growth cones of
receptor-ligand interaction radius $r_{\!_J}$ a distance $d_{kb}$ from one another,
given by
%
\begin{equation}
Q_{\!_J}(d_{kb}) = \begin{cases}
     \sum_i^N F_i\,r_{i,b}\,l_{i,k}    & d_{kb} \leq 2r_{\!_J} \\
     0 & d_{kb} > 2r_{\!_J}
     \end{cases}
\end{equation}
%
The sign of $Q_{\!_J}$ (which is dependent on the values of $F_i$) determines
whether the interaction, $\mathbf{J}_b$, is repulsive or attractive. $l_{i,k}$
is the expression of ligand $i$ on branch $k$.

@@

@@

% ./build/sim/agent/agent1 configs/simpler/m_ee_GC_best_1.json configs/simpler/e_wt_figcomp1_exit_true.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GC_best_1_wt_figcomp1_exit_true.png}
\caption{$\mathbf{GC}$ model}
\label{f:GC}
\end{figure}

@@

% No need to show the GI model
% ./build/sim/agent/agent1 configs/simpler/m_ee_GI_best_1.json configs/simpler/e_wt_figcomp2.json
%\begin{figure}
%\includegraphics[width=\linewidth]{./images/j4_ee_GI_best_1_wt_figcomp2_exit_true.png}
%\caption{$\mathbf{GI}$ model}
%\label{f:GI}
%\end{figure}

% ./scripts/fig_GI_surgical.sh
%\begin{figure}
%\includegraphics[width=\linewidth]{./images/fig_GI_surgical.png}
%\caption{$\mathbf{GI}$ model, surgical manipulations}
%\label{f:GIsurg}
%\end{figure}
